---
title: "Assumptions and Statistical Comparisons -- TRT vs CTRL"
output: html_document
date: "2024-11-11"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#Analysis overview
For each day:
    1. Compare T21vD21 within each treatment 
    2. Compare TRTvCTRL for each cell type 

## Data Inspection and Summary Stats:
STEP1: Load Tidy Data 
STEP2: Check Assumptions
STEP3: Run Statistical Analysis  
STEP4: Generate Matching Bar Plots 

```{r loading libraries, include=FALSE}

library(readr)
library(dplyr)
library(tidyverse)
library(ggpubr)
library(ggthemes)
library(ggplot2)

```

#STEP1: Load Data
```{r load data, include=FALSE}

data <- read_csv("01_tidy_data/data.csv")
long_data <- read_csv("01_tidy_data/long_data.csv")
color_map <- read_csv("01_tidy_data/color_map.csv")
measurements <- readRDS("01_tidy_data/measurements.rds")
treatment_order <- readRDS("01_tidy_data/treatment_order.rds")
sumstats_all <- read_csv("02_tables/sumstats_all.csv")

```

# Step2: Check  Assumptions for Each Flow Measurment
```{r variance and normality }
library(car) #for levenes test

# Data frame to store results
assumption_results <- data.frame(
  Day = integer(),
  Measurement = character(),
  Treatment1 = character(),
  Treatment2 = character(),
  Cell1 = character(),
  Cell2 = character(),
  Comparison = character(),
  Control_Shapiro_PValue = numeric(),
  Treatment_Shapiro_PValue = numeric(),
  Normality = character(),
  Levene_PValue = numeric(),
  Variance_equal = logical(),
  stringsAsFactors = FALSE
)

# Specify  control 
control_label <- "CTRL" 

# Loop through days and measurement to run tests
for (day_val in unique(data$day)) {
  for (measure in measurements) {
    
    # Filter data for the current day and measurement
    day_data <- data %>%
      filter(day == day_val) %>%
      select(cell, treatment, !!sym(measure)) %>%
      na.omit() 
    
    #Using day data loop through cell type to compare each treatment to control
    for (cell_type in unique(day_data$cell)) {
      cell_data <- day_data %>% filter(cell == cell_type)
      
      # Filter control
      control_data <- cell_data %>% filter(treatment == control_label)
      
      # Within each cell type loop through treatments for comparing against control
      for (treatment_val in unique(cell_data$treatment)) {
        if (treatment_val != control_label) {
          
          # Filter for treatment group 
          treatment_data <- cell_data %>% filter(treatment == treatment_val)
          
          # Combine control and treatment data
          comparison_data <- bind_rows(control_data, treatment_data)
          
          # Run tests if at least two groups (control and  treatment)
          if (n_distinct(comparison_data$treatment) >= 2) {
            
            # Shapiro-Wilk test for normality within each group
            control_shapiro <- shapiro.test(control_data[[measure]])$p.value
            treatment_shapiro <- shapiro.test(treatment_data[[measure]])$p.value
            normality <- ifelse(control_shapiro > 0.05 & treatment_shapiro > 0.05, "Normal", "Not Normal")
            
            # Levene’s test for homogeneity of variance 
            levene_test <- leveneTest(as.formula(paste(measure, "~ treatment")), data = comparison_data)
            variance_equal <- levene_test$`Pr(>F)`[1] > 0.05  # TRUE if variances are equal
            
            # Store results for normality and variance tests in a single entry
            assumption_results <- assumption_results %>%
              bind_rows(data.frame(
                Day = day_val,
                Measurement = measure,
                Treatment1 = control_label,
                Treatment2 = treatment_val,
                Cell1 = cell_type,
                Cell2 = "NA",
                Comparison = paste0(cell_type, "--", treatment_val, "vs", control_label),
                Control_Shapiro_PValue = control_shapiro,
                Treatment_Shapiro_PValue = treatment_shapiro,
                Normality = normality,
                Levene_PValue = levene_test$`Pr(>F)`[1],
                Variance_equal = variance_equal
              ))
          }
        }
      }
    }
    
    # Loop through treatments to compare cell types within that treatment
    for (treatment_val in unique(day_data$treatment)) {
      treatment_data <- day_data %>% filter(treatment == treatment_val)
      
      # Run tests if there are at least two unique cell types
      if (n_distinct(treatment_data$cell) >= 2) {
        
        # designate one cell as "control" and the other as "treatment" for storing data 
        control_cell <- "S3D21"
        treatment_cell <- "S3T21"
        
        # Filter data for the control and treatment cell types
        control_data <- treatment_data %>% filter(cell == control_cell)
        treatment_data <- treatment_data %>% filter(cell == treatment_cell)
        
        # Combine control and treatment data
        comparison_data <- bind_rows(control_data, treatment_data)
          
        control_shapiro <- shapiro.test(control_data[[measure]])$p.value
        treatment_shapiro <- shapiro.test(treatment_data[[measure]])$p.value
        normality <- ifelse(control_shapiro > 0.05 & treatment_shapiro > 0.05, "Normal", "Not Normal")
   
        levene_test <- leveneTest(as.formula(paste(measure, "~ cell")), data = comparison_data)
        variance_equal <- levene_test$`Pr(>F)`[1] > 0.05  # TRUE if variances are equal

        assumption_results <- assumption_results %>%
          bind_rows(data.frame(
            Day = day_val,
            Measurement = measure,
            Treatment1 = treatment_val,
            Treatment2 = "NA",
            Cell1 = control_cell,
            Cell2 = treatment_cell,
            Comparison = paste0(treatment_val, "--", treatment_cell, "vs", control_cell),
            Control_Shapiro_PValue = control_shapiro,  
            Normality = normality,
            Treatment_Shapiro_PValue = treatment_shapiro,
            Levene_PValue = levene_test$`Pr(>F)`[1],
            Variance_equal = variance_equal
          ))
      }
    }
  }
}

write.csv(assumption_results,"02_tables/assumption_results.csv", row.names = FALSE)
print(assumption_results)

```

# STEP3: Run Statistical Analysis
Common tests in one loop
1. Comparison of T21 vs D21 for each treatment for each day
2. Comparision of treatment vs control for each cell type for each day 

```{r comparison }
results_cell <- data.frame(
  Measurement = character(),
  Day = numeric(),
  Treatment = character(),
  Paired_TTest_PValue = numeric(),
  Welch_TTest_PValue = numeric(),
  Kruskal_Wallis_PValue = numeric(),
  Wilcoxon_PValue = numeric(),
  stringsAsFactors = FALSE
)

results_treatment <- data.frame(
  Measurement = character(),
  Day = numeric(),
  Cell = character(),
  Compared_Treatment = character(),  
  Paired_TTest_PValue = numeric(),
  Welch_TTest_PValue = numeric(),
  Kruskal_Wallis_PValue = numeric(),
  Wilcoxon_PValue = numeric(),
  stringsAsFactors = FALSE
)

# Outer loop: Compare each treatment to the control group
control_group <- "CTRL"  # Define control group 
treatment_groups <- unique(data$treatment[data$treatment != control_group])

for (treatment_val in treatment_groups) {
  # Subset data for the current treatment and control
  subset_data <- data %>%
    filter(treatment %in% c(control_group, treatment_val))

  # Inner Loop 1: Compare cell types within a treatment and day for each measurement
  for (measure in measurements) {
    for (day_val in unique(subset_data$day)) {
      for (current_treatment in unique(subset_data$treatment)) {
       
         # Filter data for current day, treatment, and measurement
        data_day_treatment <- subset_data %>%
          filter(day == day_val, treatment == current_treatment)

        # Perform tests if there are at least two cell types
        if (length(unique(data_day_treatment$cell)) > 1) {
          kruskal_test <- kruskal.test(as.formula(paste(measure, "~ cell")), data = data_day_treatment)
          
          
          # Run Wilcoxon, Welch, and Paired t-tests if exactly two cell types
          if (length(unique(data_day_treatment$cell)) == 2) {
            wilcox_test <- wilcox.test(as.formula(paste(measure, "~ cell")), data = data_day_treatment)
            wilcox_p <- wilcox_test$p.value
            
            welch_test <- t.test(as.formula(paste(measure, "~ cell")), data = data_day_treatment, var.equal = FALSE)
            welch_p <- welch_test$p.value
            
            # Paired t-test if matching subjects are found
            if (!any(duplicated(data_day_treatment$subject))) {
              paired_test <- t.test(as.formula(paste(measure, "~ cell")), data = data_day_treatment, paired = TRUE)
              paired_p <- paired_test$p.value
            }
          }
          
          # Store results for cell type comparisons
          results_cell <- rbind(results_cell, data.frame(
            Measurement = measure,
            Day = day_val,
            Treatment = current_treatment,
            Paired_TTest_PValue = paired_p,
            Welch_TTest_PValue = welch_p,
            Kruskal_Wallis_PValue = kruskal_test$p.value,
            Wilcoxon_PValue = wilcox_p
          ))
        }
      }
    }
  }
  
  # Inner Loop 2: Compare treatments within each cell type and day for each measurement
  for (measure in measurements) {
    for (day_val in unique(subset_data$day)) {
      for (cell_val in unique(subset_data$cell)) {
        
        # Filter data for current day, cell type, and measurement
        data_day_cell <- subset_data %>%
          filter(day == day_val, cell == cell_val)
        
        # Perform tests only if there are at least two treatments
        if (length(unique(data_day_cell$treatment)) > 1) {
          kruskal_test <- kruskal.test(as.formula(paste(measure, "~ treatment")), data = data_day_cell)
          
          # Run Wilcoxon, Welch, and Paired t-tests if exactly two treatments
          if (length(unique(data_day_cell$treatment)) == 2) {
            wilcox_test <- wilcox.test(as.formula(paste(measure, "~ treatment")), data = data_day_cell)
            wilcox_p <- wilcox_test$p.value
            
            welch_test <- t.test(as.formula(paste(measure, "~ treatment")), data = data_day_cell, var.equal = FALSE)
            welch_p <- welch_test$p.value
            
            # Paired t-test if matching subjects are found
            if (!any(duplicated(data_day_cell$subject))) {
              paired_test <- t.test(as.formula(paste(measure, "~ treatment")), data = data_day_cell, paired = TRUE)
              paired_p <- paired_test$p.value
            }
          }
          
          # Store results
          results_treatment <- rbind(results_treatment, data.frame(
            Measurement = measure,
            Day = day_val,
            Cell = cell_val,
            Compared_Treatment = paste(treatment_val, "vs", control_group),
            Paired_TTest_PValue = paired_p,
            Welch_TTest_PValue = welch_p,
            Kruskal_Wallis_PValue = kruskal_test$p.value,
            Wilcoxon_PValue = wilcox_p
            
          ))
        }
      }
    }
  }
}


print(results_cell)
print(results_treatment)

 
```

```{r save stats }
write.csv(results_cell,"02_tables/cell_stats.csv")
write.csv(results_treatment,"02_tables/treatment_stats.csv")
```

```{r save stats }

```

#STEP4: Generate Matching Barplots  ---
#DONT NEED HERE ONLY RA AND CTRL HERE


```{r create subfolder for individual treatment graphs}
# Define the path and folder name
folder_path <- "03_plots/vsCTRL"

# Create the folder
if (!dir.exists(folder_path)) {
  dir.create(folder_path, recursive = TRUE)
  message("Folder created at: ", folder_path)
} else {
  message("Folder already exists at: ", folder_path)
}
```

```{r vs CTRL }

# Define the control group manually
control_group <- "CTRL"


# Loop through each measurement 
for (measure in measurements) {
  
  # Filter data for  current measurement
  summary_data <- sumstats_all %>% 
    filter(measurement == measure)%>%
    na.omit()
  
  measure_data <- long_data %>%
    filter(measurement == measure)%>%
    na.omit() 
  
  
  # Loop through treatments
  for (treatment_val in unique(measure_data$treatment)) {
    if (treatment_val != control_group) {
      
      # Filter data for control and current treatment
      comparison_data <- measure_data %>% filter(treatment %in% c(control_group, treatment_val))
      comparison_summary <- summary_data %>% filter(treatment %in% c(control_group, treatment_val))
      
      # First Plot: Comparing cell types, with days on x-axis and fill by cell
      plot1 <- ggplot(comparison_summary, aes(x = factor(day), y = mean, fill = cell)) +
        geom_bar(width=.8, stat = "identity", position = position_dodge(width = 0.9), alpha = 0.8) +
        scale_y_continuous(labels = scales::percent_format(,1)) +
        geom_point(data = comparison_data, aes(x = factor(day), y = value), 
                   position = position_jitterdodge(jitter.width = 0.1, dodge.width = 0.9), 
                   size = 1.5, alpha = 0.6) +
        geom_errorbar(aes(ymin = mean - sd, ymax = mean + sd),
                      position = position_dodge(width = 0.9), width = 0.2, color = "black") +
        scale_fill_manual(values = setNames(color_map$Color, color_map$Group)) +
        facet_wrap(~ treatment) +
        labs(
          title = paste(measure, "- Comparison of Cell Types in", treatment_val, "&", control_group),
          x = "Day",
          y = measure) +
        theme_bw() +
        theme(legend.position = "top",
          legend.text = element_text(size = 14),
          axis.text.x = element_text(size = 12), 
          axis.text.y = element_text(size = 11),
          axis.title.x = element_text(size = 14),
          axis.title.y = element_text(size = 14))
      
      #Define filename
      file_name1 <- paste0(measure, "--", treatment_val, "vs", control_group,  "_facet_treatment.pdf")
      
      ggsave(file_name1, path = "03_plots/vsCTRL", plot = plot1, width = 8, height = 6, units = "in")
      
      # Second Plot: Comparing treatments, with days on x-axis and fill by treatment
      plot2 <- ggplot(comparison_summary, aes(x = factor(day), y = mean, fill = treatment)) +
        geom_bar(width=.8, stat = "identity", position = position_dodge(width = 0.9), alpha = 0.8) +
        scale_y_continuous(labels = scales::percent_format(,1)) +
        geom_point(data = comparison_data, aes(x = factor(day), y = value), 
                   position = position_jitterdodge(jitter.width = 0.1, dodge.width = 0.9), 
                   size = 1.5, alpha = 0.6) +
        geom_errorbar(aes(ymin = mean - sd, ymax = mean + sd),
                      position = position_dodge(width = 0.9), width = 0.2, color = "black") +
        scale_fill_manual(values = setNames(color_map$Color, color_map$Group)) +
        facet_wrap(~ cell) +
        labs(
          title = paste(measure, "- Comparison of", treatment_val, "vs", control_group),
          x = "Day",
          y = measure) +
        theme_bw() +
        theme(legend.position = "top",
          legend.text = element_text(size = 14),
          axis.text.x = element_text(size = 12), 
          axis.text.y = element_text(size = 11),
          axis.title.x = element_text(size = 14),
          axis.title.y = element_text(size = 14))
      
      file_name2 <- paste0(measure, "--", treatment_val, "vs", control_group, "_facet_cell.pdf")
      
      ggsave(file_name2, path = "03_plots/vsCTRL", plot = plot2, width = 8, height = 6, units = "in")
    }
  }
}
 
```


Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
